### 2021-12-03 lmcs/introduction.tex
Update introduction.tex

\section{Introduction}
Session types~\cite{honda93,takeuchihonda94,hondavasconcelos98} are types for protocols. Regular types ensure functions are used according to their specification. Session types ensure \emph{communication channels} are used according to their protocols. Session types have been studied in many settings. For instance, in the $\pi$-calculus~\cite{honda93,takeuchihonda94,hondavasconcelos98}, a foundational calculus for communication and concurrency, and in concurrent $\lambda$-calculi~\cite{gayvasconcelos10}, including the focus of our paper: Good Variation~\cite[GV]{wadler15,lindleymorris15}.

@@

Our aim is to develop a more expressive version of GV while maintaining deadlock freedom. While process calculi have their advantages, \eg, their succinctness, we chose to work with GV for several reasons. In general, concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. Case in point: we easily adapted the type system of PGV to Linear Haskell~\cite{bernardyboespflug18}, which gives us a library for deadlock-free session-typed programming~\cite{kokkedardha21hs}.
The benefit of working specifically with GV, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its formal properties, including deadlock freedom. We thus pose our research question for GV:
\begin{quotation}
  \textbf{RQ:}
  Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?
\end{quotation}

We follow the line of work from CP to Priority CP, and present Priority GV (PGV), a~variant of GV which decouples channel creation from thread spawning, thus allowing cyclic-structured processes, but which nonetheless guarantees deadlock freedom via priorities. This closes the circle of the connection between CP and GV~\cite{wadler14}, and their priority-based versions, PCP~\cite{dardhagay18extended} and PGV.
We make the following main contributions:
\begin{enumerate}[labelindent=0pt,labelwidth=1.333\parindent,align=left,labelsep*=0pt,leftmargin=!]
\item[(\secref{sec:pgv})] \textbf{Priority GV}. We present Priority GV (\secref{sec:pgv}, PGV), a session-typed functional language with priorities, and prove subject reduction (\cref{thm:pgv-subject-reduction-confs}) and progress (\cref{thm:pgv-closed-progress-confs}). We addresses several problems in the original GV language, most notably:
  \begin{enumerate*}
  \item PGV does not require the pseudo-type $S^\sharp$; and
  \item its structural congruence is type preserving.
  \end{enumerate*}
  PGV answers our research question positively as it allows cyclic-structured binary session-typed processes that are deadlock free.
\item[(\secref{sec:pcp})] \textbf{Translation from PCP to PGV}.
  We present a \emph{sound and complete encoding} of Priority CP~\cite{dardhagay18extended} in PGV (\secref{sec:pcp}). We prove the encoding preserves typing (\cref{thm:pcp-to-pgv-confs-preservation}) and satisfies operational correspondence (\cref{thm:pcp-to-pgv-operational-correspondence-soundness,thm:pcp-to-pgv-operational-correspondence-completeness}).
  To obtain a tight correspondence, we update PCP, moving away from commuting conversions and reduction as cut elimination towards reduction based on structural congruence, as it is standard in process calculi.
\end{enumerate}

>>>>

\section{Introduction}
Session types~\cite{honda93,takeuchihonda94,hondavasconcelos98} are a type formalism used to specify and verify correct ordering of operations, namely protocols. Regular types ensure, for example that functions are used according to their specification. Session types ensure \emph{communication channels} are used according to their protocols. Session types have been studied in many settings. Most notably, they have been defined for the $\pi$-calculus~\cite{honda93,takeuchihonda94,hondavasconcelos98}, a foundational calculus for communication and concurrency, and the concurrent $\lambda$-calculi~\cite{gayvasconcelos10}, including the main focus of our paper: Good Variation~\cite[GV]{wadler15,lindleymorris15}.

@@

Our aim is to develop a more expressive version of GV while maintaining deadlock freedom. While process calculi have their advantages, \eg, their succinctness, we chose to work with GV for several reasons. In general, concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. Case in point: we easily adapted the type system of PGV to Linear Haskell~\cite{bernardyboespflug18}, which gives us a library for deadlock-free session-typed programming~\cite{kokkedardha21hs}.
The benefit of working specifically with GV, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its formal properties, including deadlock freedom.

We thus pose our research question for GV:

\begin{quotation}
  \textbf{RQ:}
  Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?
\end{quotation}

We follow the line of work from CP to Priority CP, and present Priority GV (PGV), a~variant of GV which decouples channel creation from thread spawning, thus allowing cyclic-structured processes, but which nonetheless guarantees deadlock freedom via priorities. This closes the circle of the connection between CP and GV~\cite{wadler14}, and their priority-based versions, PCP~\cite{dardhagay18extended} and PGV.
We make the following main contributions:
\begin{enumerate}[labelindent=0pt,labelwidth=1.333\parindent,align=left,labelsep*=0pt,leftmargin=!]
\item[(\cref{sec:pgv})] \textbf{Priority GV}. We present Priority GV (\cref{sec:pgv}, PGV), a session-typed functional language with priorities, and prove \emph{subject reduction} (\cref{thm:pgv-subject-reduction-confs}) and \emph{progress} (\cref{thm:pgv-closed-progress-confs}). We addresses several problems in the original GV language, most notably:
  \begin{enumerate}
  \item PGV does not require the pseudo-type $S^\sharp$; and
  \item its structural congruence is type preserving.
  \end{enumerate}
  PGV answers our research question positively as it allows cyclic-structured binary session-typed processes that are deadlock free.
\item[(\cref{sec:pcp})] \textbf{Translation from PCP to PGV}.
  We present a \emph{sound and complete encoding} of PCP~\cite{dardhagay18extended} in PGV (\cref{sec:pcp}). We prove the encoding preserves typing (\cref{thm:pcp-to-pgv-confs-preservation}) and satisfies operational correspondence (\cref{thm:pcp-to-pgv-operational-correspondence-soundness,thm:pcp-to-pgv-operational-correspondence-completeness}).
  To obtain a tight correspondence, we update PCP, moving away from commuting conversions and reduction as cut elimination towards reduction based on structural congruence, as it is standard in process calculi.
\end{enumerate}
